\section{CVS}

The CVS repository resides in \C{/user/form/cvs\_repository}. It is advisable to
set the enviroment variable \C{CVSROOT} accordingly, like (using bash shell syntax)
{\scriptsize
\begin{verbatim}
  export CVSROOT=:ext:myusername@mytrustedmachine.nikhef.nl:/user/form/cvs_repository
\end{verbatim}}

A mailing list exists for CVS commits. The administration interface for this mailing list can be
found under the web address

\LINK{https://mailman.nikhef.nl/cgi-bin/admin/form-cvs}

A password is required. 

Click {\it Membership Management} and then {\it Mass Subscription} to add
new people. The personal details of the subscribers like the email address or
the name can be changed under {\it Membership Management} as well.

The triggering of the CVS commits mails is done in the following way. In the
file \C{loginfo} in the directory \C{CVSROOT} (inside the repository) the
default action for logging is set such that the script \C{/user/form/cvs-log.sh}
will be called with the committer's user name and the CVS mailing list user
name. The shell script does some simple message transformation and then uses the
command \C{mail} to send the commit mail to the mailing list.

\subsection{Some useful CVS idioms}

To just show what would be updated/changed without actually
modifiying anything, use
\begin{verbatim}
  cvs -n update
\end{verbatim}

If \C{cvs -n update} has shown you that something new in the repository will be
merged into your directory and you want to know in advance what the details are,
you can do for each of the files involved a
\begin{verbatim}
  cvs status <filename>
\end{verbatim}
and note the version number of your local file, and then do a
\begin{verbatim}
  cvs diff -r <versionnumber> <filename>
\end{verbatim}
to see the differences.

In case you want to compile an older version of \FORM\ (maybe to find out
whether a certain bug is already present or not), do
\begin{verbatim}
  cvs update -D "<DATE>"
\end{verbatim}
to checkout the sources as they were on a certain date, e.g. \\
\C{cvs update -D "2006-05-12"}. The files will get the so-called sticky flag,
which do prevent simple \C{cvs update} commands in the future to update to the
latest version from the repository. To remove the sticky flag on a file use
\begin{verbatim}
  cvs update -A <filename>
\end{verbatim}
Without the filename all files will have the sticky flag removed.

